---
title: "Team project: Checkpoint"
author: "Juan Pablo Jaramillo & Varun Vijayakumar - TEAM 2"
subtitle: "R for Data Science @ Hult University"
output:
  html_notebook:
    highlight: pygments
    number_sections: yes
    theme: readable
    toc: yes
    toc_float:
      collapsed: yes
---
ATTENTION!!
PLEASE RUN ALL THE CODE BELOW TO ENSURE ALL THE VARIABLES ARE IN THE GLOBAL ENVIRONMENT BEFORE LOOKING THROUGH THE 'FINAL PROJECT' FILE.
SHORTCUT TO RUN ALL THE BELOW CODE: COMMAND + OPTION + R

# Set-up {.unnumbered}
Load Tidyverse
```{r load packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
# load other packages AS NEEDED and ONLY IF YOU *REALLY* NEED THEM
```

#Load Data
Load your data here:
```{r load data, message=FALSE, warning=FALSE}
fifa_data <- read.csv('data/fifa.raw.data.csv') #Import function used
```

#Exploratory data analysis
**Figure 1:** 
Histogram of Overall Scores of players.
```{r}
fifa_data %>% 
  ggplot(aes(x = Overall))+   
  geom_histogram(bins = 100)
```
**Figure 2:**
Scatter plot between Price and Overall score
```{r}
fifa_data %>% 
  filter(Price > 0) %>% 
  ggplot(data = .,aes(x = Overall, y= Price))+
  geom_point()+
  geom_vline(xintercept = 65, color = "Red")
```
**Figure 2.1**
Scatter plot between log_price and Overall score
```{r}
#TRANSFORMATION OF PRICE
fifa_data <- fifa_data %>% 
  mutate(log_price = log(Price))
```

```{r}
#Scatter plot after transformation
fifa_data %>% 
  filter(Price > 0) %>% 
  ggplot(data = .,aes(x = Overall, y= log_price))+
  geom_point()+
  geom_vline(xintercept = 65, color = "Red")+
  geom_smooth(method = "lm")
```
**Figure 3:**
Linear Regression summary statistics
```{r}
fifa_data %>% 
  filter(Price > 0) %>% 
  lm(formula = log_price ~ Overall, data = .) %>% 
  summary()
```
**Figure 4:** 
Scatter plot showing the average price per playing position
```{r, message=FALSE, warning=FALSE}
fifa_data %>% 
  group_by(Position) %>% 
  summarize(avg_price_position = mean(Price)) %>% 
  ggplot(data = ., aes(x = Position, y = avg_price_position))+
  geom_point()
```
**Figure 5:**
Outlier Identification
```{r}
fifa_data %>% 
  slice_max(Price, n=30) %>% 
  arrange(desc(Price))
```
**Figure 6:**
Analysis of relationship between salary and price of players
```{r}
fifa_data %>% 
  ggplot(data = ., aes(x = Price, y = Salary))+
  geom_point(alpha = 0.25)+
  geom_vline(xintercept = 30000000, color = "red")+
  geom_smooth(method = "lm")
```

# Analysis plan
##Player grouping
```{r}
fifa_data <- fifa_data %>% 
  mutate(Grouped_Position = ifelse(fifa_data$Position == 'RF', 'STRIKER',
                        ifelse(fifa_data$Position == 'CAM', 'MIDFIELD',
                        ifelse(fifa_data$Position == 'CB', 'DEFENCE',
                        ifelse(fifa_data$Position == 'CDM', 'MIDFIELD',
                        ifelse(fifa_data$Position == 'CF', 'STRIKER',
                        ifelse(fifa_data$Position == 'CM', 'MIDFIELD',
                        ifelse(fifa_data$Position == 'GK', 'GOAL-KEEPER',
                        ifelse(fifa_data$Position == 'LAM', 'MIDFIELD',
                        ifelse(fifa_data$Position == 'LB', 'DEFENCE',
                        ifelse(fifa_data$Position == 'LCB', 'DEFENCE',
                        ifelse(fifa_data$Position == 'LDM', 'MIDFIELD',
                        ifelse(fifa_data$Position == 'LCM', 'MIDFIELD',
                        ifelse(fifa_data$Position == 'LF', 'WING',
                        ifelse(fifa_data$Position == 'LM', 'MIDFIELD',
                        ifelse(fifa_data$Position == 'LW', 'WING',
                        ifelse(fifa_data$Position == 'LWB', 'DEFENCE', 
                        ifelse(fifa_data$Position == 'RAM', 'MIDFIELD', 
                        ifelse(fifa_data$Position == 'RB', 'DEFENCE', 
                        ifelse(fifa_data$Position == 'RCB', 'DEFENCE',
                        ifelse(fifa_data$Position == 'RCM', 'MIDFIELD',
                        ifelse(fifa_data$Position == 'RDM', 'MIDFIELD',
                        ifelse(fifa_data$Position == 'RF', 'STRIKER',
                        ifelse(fifa_data$Position == 'RM', 'MIDFIELD',
                        ifelse(fifa_data$Position == 'RS', 'STRIKER',
                        ifelse(fifa_data$Position == 'RW', 'WING',
                        ifelse(fifa_data$Position == 'RWB', 'DEFENCE',
                        ifelse(fifa_data$Position == 'ST', 'STRIKER',
                        ifelse(fifa_data$Position == 'LS', 'STRIKER',
                        'other')))))))))))))))))))))))))))))
```

```{r}
unique(fifa_data$Grouped_Position)
```

##Price analysis
```{r}
fifa_data %>% # Create a boxplot to understand the outliers (Superstars)
  ggplot(data = ., mapping = aes(x= log_price)) +
  geom_boxplot()
```
```{r}
#Calculate the range 
range(fifa_data$Price)
```

```{r}
#Calculate the Quantiles to see the the distribution of prices under the curve
quantile(fifa_data$Price)
```

```{r} 
# Filter data set by price, international reputation, and Potential to create the avg player data set
fifa_data_avg_player <- fifa_data %>% 
  filter(.data = ., Price < 9000000) 
```

```{r}
#check the range in prices of the avg players
range(fifa_data_avg_player$Price)
```

```{r}
#look at the distribution in prices of the avg player 
quantile(fifa_data_avg_player$Price) 
```

```{r}
fifa_superstars <- fifa_data %>% 
  filter(., Price > 9000000)  #number of superstar that we are not considering that are valued over $9,000,000 i.e SuperStars
```

```{r}
fifa_data_avg_player %>% 
  group_by(Grouped_Position) %>% 
  count() 
```
##Superstar Averages
```{r superstar averages, message=FALSE, warning=FALSE}
superstar_avg = fifa_superstars %>% 
  group_by(Grouped_Position) %>% 
  summarise(across(everything(), mean))
```


##Identifying stastically significant skills per position
**Playing position: STRIKER**
```{r}
#Finding out important attributes for Strikers players
fifa_data_avg_player %>% 
  filter(Price > 0 & Grouped_Position == "STRIKER") %>%
  lm(formula = log_price ~ Crossing+Finishing+HeadingAccuracy+ShortPassing+Volleys+Dribbling+Curve+FKAccuracy+LongPassing+BallControl+Acceleration+SprintSpeed+Agility+Reactions+Balance+ShotPower+Jumping+Stamina+Strength+LongShots+Aggression+Interceptions+Positioning+Vision+Penalties+Composure+Marking+StandingTackle+SlidingTackle+GKDiving+GKHandling+GKKicking+GKReflexes, data = .) %>% 
  summary()
```
**Playing position:WING**
```{r}
#Finding out important attributes for Wing players
fifa_data_avg_player %>% 
  filter(Price > 0 & Grouped_Position == "WING") %>%
  lm(formula = log_price ~ Crossing+Finishing+HeadingAccuracy+ShortPassing+Volleys+Dribbling+Curve+FKAccuracy+LongPassing+BallControl+Acceleration+SprintSpeed+Agility+Reactions+Balance+ShotPower+Jumping+Stamina+Strength+LongShots+Aggression+Interceptions+Positioning+Vision+Penalties+Composure+Marking+StandingTackle+SlidingTackle+GKDiving+GKHandling+GKKicking+GKReflexes, data = .) %>% 
  summary()
```
**Playing position:MIDFIELD**
```{r}
#filtering for MIDFIELD players
fifa_data_avg_player %>% 
  filter(Price > 0 & Grouped_Position == "MIDFIELD") %>%
  lm(formula = log_price ~ Crossing+Finishing+HeadingAccuracy+ShortPassing+Volleys+Dribbling+Curve+FKAccuracy+LongPassing+BallControl+Acceleration+SprintSpeed+Agility+Reactions+Balance+ShotPower+Jumping+Stamina+Strength+LongShots+Aggression+Interceptions+Positioning+Vision+Penalties+Composure+Marking+StandingTackle+SlidingTackle+GKDiving+GKHandling+GKKicking+GKReflexes, data = .) %>% 
  summary()

```
**Playing position:DEFENCE**
```{r}
#filtering for DEFENCE players
fifa_data_avg_player %>% 
  filter(Price > 0 & Grouped_Position == "DEFENCE") %>%
  lm(formula = log_price ~ Crossing+Finishing+HeadingAccuracy+ShortPassing+Volleys+Dribbling+Curve+FKAccuracy+LongPassing+BallControl+Acceleration+SprintSpeed+Agility+Reactions+Balance+ShotPower+Jumping+Stamina+Strength+LongShots+Aggression+Interceptions+Positioning+Vision+Penalties+Composure+Marking+StandingTackle+SlidingTackle+GKDiving+GKHandling+GKKicking+GKReflexes, data = .) %>% 
  summary()
```

**Playing position:GOAL-KEEPER**
```{r}
#filtering for GOAL-KEEPER players
fifa_data_avg_player %>% 
  filter(Price > 0 & Grouped_Position == "GOAL-KEEPER") %>%
  lm(formula = log_price ~ Crossing+Finishing+HeadingAccuracy+ShortPassing+Volleys+Dribbling+Curve+FKAccuracy+LongPassing+BallControl+Acceleration+SprintSpeed+Agility+Reactions+Balance+ShotPower+Jumping+Stamina+Strength+LongShots+Aggression+Interceptions+Positioning+Vision+Penalties+Composure+Marking+StandingTackle+SlidingTackle+GKDiving+GKHandling+GKKicking+GKReflexes, data = .) %>% 
  summary()
```

##Player Identification
###FILTERING FOR STRIKERS - 2 Main SKILLS
```{r}
fifa_data_avg_player %>%
  filter(Grouped_Position == "STRIKER") %>% 
  filter(Finishing >= 80.70 & HeadingAccuracy >= 73.98) %>% 
  arrange(desc(Price))
```


###FILTERING FOR WING - 2 Main SKILLS
```{r}
fifa_data_avg_player %>%
  filter(Grouped_Position == "WING") %>% 
  filter(SprintSpeed >= 82.84 & Crossing >= 75.82) %>% 
  arrange(desc(Price))
```

###Filtering for midfield - 2 main SKILLS
```{r}
fifa_data_avg_player %>% 
  filter(Grouped_Position == "MIDFIELD") %>%   
  filter(ShortPassing >= 79.67 & BallControl >= 80.84) %>% 
  arrange(desc(Price))
```

###FILTERING FOR DEFENCE - 2 Main SKILLS
```{r}
fifa_data_avg_player %>% 
  filter(Grouped_Position == "DEFENCE") %>%   
  filter(Strength >= 77.35 & SlidingTackle >= 79.7 & StandingTackle >= 81) %>% 
  arrange(desc(Price))
```
###FILTERING FOR Goal-keeper - 2 Main SKILLS
```{r}
fifa_data_avg_player %>% 
  filter(Grouped_Position == "GOAL-KEEPER") %>% 
  filter(GKDiving >= 82.34 & GKReflexes > 84.01 ) %>%
  arrange(desc(Price))
```

###Predicting Prices (Strikers) with Linear regression formula
```{r}
lm_strikers = fifa_data_avg_player %>% 
  filter(Price > 0 & Grouped_Position == "STRIKER") %>%
  lm(formula = log_price ~ Crossing+Finishing+HeadingAccuracy+ShortPassing+Volleys+Dribbling+Curve+FKAccuracy+LongPassing+BallControl+Acceleration+SprintSpeed+Agility+Reactions+Balance+ShotPower+Jumping+Stamina+Strength+LongShots+Aggression+Interceptions+Positioning+Vision+Penalties+Composure+Marking+StandingTackle+SlidingTackle+GKDiving+GKHandling+GKKicking+GKReflexes, data = .)
  

strikers_filtered = fifa_data_avg_player %>%
  filter(Grouped_Position == "STRIKER" & Price > 0)

predict_p_df = as_data_frame(broom::augment(lm_strikers))

strikers_predited <- cbind(strikers_filtered, predict_p_df$.fitted)

strikers_predited %>%
  filter(`predict_p_df$.fitted`< log_price)


```

###Predicting Prices (Wing) with Linear regression formula
```{r}
lm_wing = fifa_data_avg_player %>% 
  filter(Price > 0 & Grouped_Position == "WING") %>%
  lm(formula = log_price ~ Crossing+Finishing+HeadingAccuracy+ShortPassing+Volleys+Dribbling+Curve+FKAccuracy+LongPassing+BallControl+Acceleration+SprintSpeed+Agility+Reactions+Balance+ShotPower+Jumping+Stamina+Strength+LongShots+Aggression+Interceptions+Positioning+Vision+Penalties+Composure+Marking+StandingTackle+SlidingTackle+GKDiving+GKHandling+GKKicking+GKReflexes, data = .)
  

wing_filtered = fifa_data_avg_player %>%
  filter(Grouped_Position == "WING" & Price > 0)

predict_wing_df = as_data_frame(broom::augment(lm_wing))

wing_predited <- cbind(wing_filtered, predict_wing_df$.fitted)

wing_predited %>%
  filter(`predict_wing_df$.fitted`< log_price)
```

###Predicting Prices (MIDFIELD) with Linear regression formula
```{r}
lm_midfield = fifa_data_avg_player %>% 
  filter(Price > 0 & Grouped_Position == "MIDFIELD") %>%
  lm(formula = log_price ~ Crossing+Finishing+HeadingAccuracy+ShortPassing+Volleys+Dribbling+Curve+FKAccuracy+LongPassing+BallControl+Acceleration+SprintSpeed+Agility+Reactions+Balance+ShotPower+Jumping+Stamina+Strength+LongShots+Aggression+Interceptions+Positioning+Vision+Penalties+Composure+Marking+StandingTackle+SlidingTackle+GKDiving+GKHandling+GKKicking+GKReflexes, data = .)
  

midfield_filtered = fifa_data_avg_player %>%
  filter(Grouped_Position == "MIDFIELD" & Price > 0)

predict_midfield_df = as_data_frame(broom::augment(lm_midfield))

midfield_predited <- cbind(midfield_filtered, predict_midfield_df$.fitted)

midfield_predited %>%
  filter(`predict_midfield_df$.fitted`< log_price)
```

###Predicting Prices (DEFENCE) with Linear regression formula
```{r}
lm_defence = fifa_data_avg_player %>% 
  filter(Price > 0 & Grouped_Position == "DEFENCE") %>%
  lm(formula = log_price ~ Crossing+Finishing+HeadingAccuracy+ShortPassing+Volleys+Dribbling+Curve+FKAccuracy+LongPassing+BallControl+Acceleration+SprintSpeed+Agility+Reactions+Balance+ShotPower+Jumping+Stamina+Strength+LongShots+Aggression+Interceptions+Positioning+Vision+Penalties+Composure+Marking+StandingTackle+SlidingTackle+GKDiving+GKHandling+GKKicking+GKReflexes, data = .)
  

defence_filtered = fifa_data_avg_player %>%
  filter(Grouped_Position == "DEFENCE" & Price > 0)

predict_defence_df = as_data_frame(broom::augment(lm_defence))

defence_predited <- cbind(defence_filtered, predict_defence_df$.fitted)

defence_predited %>%
  filter(`predict_defence_df$.fitted`< log_price)
```

###Predicting Prices (GOAL-KEEPER) with Linear regression formula
```{r}
lm_goalie = fifa_data_avg_player %>% 
  filter(Price > 0 & Grouped_Position == "GOAL-KEEPER") %>%
  lm(formula = log_price ~ Crossing+Finishing+HeadingAccuracy+ShortPassing+Volleys+Dribbling+Curve+FKAccuracy+LongPassing+BallControl+Acceleration+SprintSpeed+Agility+Reactions+Balance+ShotPower+Jumping+Stamina+Strength+LongShots+Aggression+Interceptions+Positioning+Vision+Penalties+Composure+Marking+StandingTackle+SlidingTackle+GKDiving+GKHandling+GKKicking+GKReflexes, data = .)
  

goalie_filtered = fifa_data_avg_player %>%
  filter(Grouped_Position == "GOAL-KEEPER" & Price > 0)

predict_goalie_df = as_data_frame(broom::augment(lm_goalie))

goalie_predited <- cbind(goalie_filtered, predict_goalie_df$.fitted)

goalie_predited %>%
  filter(`predict_goalie_df$.fitted`< log_price)
```

#Output
Selecting Top Players for MIDFLIED Position, message=FALSE, warning=FALSE
```{r}
midfield_predited %>% 
  filter(ShortPassing >= 79.67 & BallControl >= 80.84 & `predict_midfield_df$.fitted`> log_price) %>% 
  arrange(desc(Price))
```

Selecting Top Players for GOAL - KEEPER Position, message=FALSE, warning=FALSE
```{r}
goalie_predited %>% 
  filter(GKDiving >= 82.34 & GKReflexes > 84.01  & `predict_goalie_df$.fitted`> log_price) %>% 
  arrange(desc(Price))
```

Selecting Top Players for DEFENCE Position, message=FALSE, warning=FALSE
```{r}
defence_predited %>% 
  filter(Strength >= 77.35 & SlidingTackle >= 79.7 & StandingTackle >= 81  & `predict_defence_df$.fitted`> log_price) %>% 
  arrange(desc(Price))
```

Selecting Top Players for WING Position, message=FALSE, warning=FALSE
```{r}
wing_predited %>% 
  filter(SprintSpeed >= 82.84 & Crossing >= 75.82  & `predict_wing_df$.fitted` > log_price) %>% 
  arrange(desc(Price))
```

Selecting Top Players for STRICKERS Position, message=FALSE, warning=FALSE
```{r}
strikers_predited %>% 
  filter(Finishing >= 80.70 & HeadingAccuracy >= 73.98  & `predict_p_df$.fitted`> log_price) %>% 
  arrange(desc(Price))
```